---
title: "Analyzing Sustainability Indicators Amidst Climate Change"
subtitle: "A deeper analysis of CO₂"
author: 
  - "Shamil Khusnullin"
---

**Overview** 

This page allows you to dive deeper into the exploration of CO₂ levels change happening over the last 30 years around the world. You can explore the CO₂ level changes on the map year by year. Furthermore, this page includes the clustering and a trained algorithm that predicts the CO₂ level change relative to 1990 based on the electricity production input.

```{r}
#| label: setup
#| include: false

# Improve digit display and NA display for kable tables
options(scipen = 1, knitr.kable.NA = "")

# Load necessary packages
if (!require('pacman')) install.packages('pacman'); library('pacman')
pacman::p_load(viridis, tidyverse, ggplot2, lubridate, knitr, dplyr, purrr, 
               broom, sf, rnaturalearth, janitor, countrycode, stringr, 
               GGally, tidyr, kableExtra, shinylive, randomForest)
load("Data/data.RData")
# Create a long CO2 df version:
co2_map <- data |> 
  filter(`Indicator Name` == 
"Carbon dioxide (CO2) emissions (total) excluding LULUCF (% change from 1990)")|> 
  select(1:36, geometry) |> 
  mutate(across(matches("^(19|20)\\d{2}$"), as.numeric)) |>
  #Pivot longer: create one row per Year and Value
  pivot_longer(!c(`Country Name`, `Country Code`, `Indicator Name`, geometry), 
               names_to  = "Year",
               values_to = "co2conc" ) |> 
  # Cap co2conc at 500 % to avoid outliers
  mutate(co2conc = pmin(co2conc, 500)) |> 
  mutate(Year = as.numeric(Year)) 
  
```


## Interactive Map
The change in atmospheric CO₂ levels is a gradual process that happens over decades. Over the past 30 years, global CO₂ emissions have surged dramatically. Remarkably, more carbon dioxide has been emitted since 1990 than in all prior human history [combined](https://thecorrespondent.com/751/weve-emitted-more-co2-in-the-past-30-years-than-in-all-of-history-these-three-reasons-are-to-blame?). To better understand the change, we’ve created an interactive map that lets you explore how CO₂ concentrations have shifted year by year, starting from the baseline year of 1991. By moving through the timeline, you can see which regions have experienced the most significant increases or decreases in emissions. 


<small>Use the slider below the map to select a specific year and observe the changes in CO₂ emissions compared to previous years. This allows you to closely examine how individual countries have increased or decreased their emissions over time. If you want full automated visualization: Click the play button to animate the map and watch the year-by-year evolution of global CO₂ emissions. </small>

<iframe src="https://shamchikkhus.shinyapps.io/MapApp/"
        width="100%" height="600px" frameborder="0">
</iframe>


## Classifying CO₂ changes
Next, we clustered the changes of CO₂ starting from 1991 up to 2015. Clustering on just one variable (percent change in CO₂ relative to 1990) makes the analysis very clear and direct and eliminates extra bias. By letting a clustering algorithm find its own group boundaries, we avoid guessing where levels (such as “moderate” or “extreme”) really begin. Instead, the data itself creates four natural clusters:

* Cluster 1: decrease or slight increase (<23.92% increase)


* Cluster 2: moderate increase (up to 119.4%)


* Cluster 3: high increase (up to 255%)


* Cluster 4: extreme increase (above 255%)


The plotted graph below should represent how these clusters looked like in 2015. You can compare this graph with the interactive map, choosing the year 2015, and gain a deeper comprehension on how the data is clustered.

```{r}
# Seed for Reproducibility
set.seed(631)

# Remove geometry and delete NA values
co2_cleaned <- co2_map |> 
  drop_na() |> 
  select(`Country Code`, `Country Name`, `Year`, `co2conc`, geometry)
  
# Choose only co2conc
co2_numeric <- co2_cleaned |> 
  select(co2conc)

# Normalize the data
co2_scaled <- scale(co2_numeric) 

# Clustering
kmeans_result <- kmeans(co2_scaled, centers = 4, nstart = 25)


# Add cluster back to co2_cleaned df
co2_cleaned$Cluster <- as.factor(kmeans_result$cluster)

co2_final <- co2_cleaned |> 
  right_join(climate_long, by =  c("Country Name", "Country Code", "Year")) |> 
  drop_na() 

co2_final <- co2_final |> 
  mutate(Cluster = case_when(
    Cluster == "1" ~ "Extreme Increase",
    Cluster == "2" ~ "Decrease/Slight Increase",
    Cluster == "3" ~ "Moderate Increase",
    Cluster == "4" ~ "High Increase"
  )) 

# Set Clusters as factor
co2_final <- co2_final |> 
  mutate(Cluster = as.factor(Cluster))

```

```{r, warning=FALSE, fig.width=8, fig.height=4}

#Filter Year 2015

co2_2015 <- co2_final |> 
  filter(Year == 2015)


# Custom color mapping
cluster_colors <- c(
  "Extreme Increase"  = "#d73027",  
  "High Increase"     = "#fc8d59",
  "Moderate Increase" = "#fee08b",  
  "Decrease/Slight Increase" = "#1a9850" 
)

co2_2015 <- st_as_sf(co2_2015)
# Plot
ggplot(co2_2015) +
  geom_sf(aes(fill = Cluster), color = "grey40", linewidth = 0.2) +
  coord_sf(crs = "+proj=robin", expand = FALSE) +
  scale_fill_manual(values = cluster_colors)+
  theme_void(base_size = 14) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 8)
  ) +
  labs(title = "2015 CO2 emission Change relative to 1990")

```

```{r}
set.seed(12)

# Train data on values from 1990 to 2014
train_data <- co2_final[co2_final$Year != 2015, ]


# Test data based on 2015
test_data <- co2_final[co2_final$Year == 2015,] 

# Build a Forest learning algorithm
# Use 4 different variables to predict a cluster
rf_model <- randomForest(Cluster ~ Renew_Cons + Renew_Elec + Oil_Elec + Gas_Elec
                         + Coal_Elec,
                         data = train_data, ntree = 100)

# Draw predictions for 2015 based on the model.
predictions <- predict(rf_model, test_data)

```
## Predicting CO₂ based on electricity production variables.

We trained a supervised random forest model to predict each country’s CO₂ level change cluster using five predictors: renewable consumption, renewable electricity output, oil‐based electricity, gas‐based electricity, and coal‐based electricity. For training dataset, we used the long CO₂ level change data set covering yearly values from 1991 through 2014. For the actual algorithm we used the supervised machine learning algorithm [randomForest](https://cran.r-project.org/web/packages/randomForest/randomForest.pdf) (number of trees = 100)

To evaluate performance, we held out the 2015 observations entirely and then asked the trained model to predict each country’s 2015 cluster. The resulting confusion matrix (shown below) compares actual versus predicted clusters: rows are the true cluster, columns are the model’s guess, and the diagonal entries are correct classifications.

```{r}
# Confusion Matrix
confusion_matrix <- table(Actual = test_data$Cluster, Predicted = predictions)
# Kable package to print the Matrix
kable(confusion_matrix, caption = "Confusion Matrix (Predicted vs Actual)")

```

We calculated overall accuracy by dividing number of correct guesses by total number of classification. The predicted accuracy was 0.725, meaning the model correctly placed about 72.5% of countries into their true 2015 cluster. Most decreases/slight increases and high‐increase cases were recovered well, though a few countries were confused between neighboring categories.

Of course, this approach has limitations. By focusing on only five energy variables, the model misses other drivers of CO₂ change-economic growth, the neigbhoring contries contributing to CO₂ change, or policy interventions. For example Ethiopia illustrates this: despite nearly 100% renewable consumption and electricity generation, its CO₂ levels still rose, so the model misclassified it. This counterintuitive result reminds us that no predictor set is perfect, and further work could include additional socioeconomic variables to capture the full complexity of emissions trends.

```{r}
# Calculate Accuracy (sum of diagonal entries over total number of entries)
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
print(paste("Prediction Accuracy:", round(accuracy, 3)))

```

You can test the alogorithm below:

<iframe src="https://shamchikkhus.shinyapps.io/PredictionApp/"
        width="100%" height="700px" frameborder="0">
</iframe>

<small> Disclaimer: this algorithm is only 72.5% accurate and can produce inaccurate results </small>

## Discussion

The results of our CO₂ change prediction model indeed show that energy usage variables can be used to predict CO₂ levels. The algorithm highlights both the potential and the limitations of using energy usage variables as indicators. While our model achieved a reasonable accuracy of 72.5%, correctly identifying many countries’ CO₂ change clusters, certain cases revealed the complexity of emissions dynamics. 

Counties such as Ethiopoia do not follow the general expected trens pointing to the need for broader, multi-dimensional models that integrate socioeconomic and policy-related variables to better capture the full picture of CO₂ change across nations.

## References
“Randomforest: Breiman and Cutlers Random Forests for Classification and Regression.” The Comprehensive R Archive Network, Comprehensive R Archive Network (CRAN), 22 Sept. 2024, [https://cran.r-project.org/web/packages/randomForest/randomForest.pdf](https://cran.r-project.org/web/packages/randomForest/randomForest.pdf). 
Holthaus, Eric. "We've Emitted More CO₂ in the Past 30 Years Than in All of History. These Three Reasons Are to Blame." The Correspondent, 2019 [https://thecorrespondent.com/751/weve-emitted-more-co2-in-the-past-30-years-than-in-all-of-history-these-three-reasons-are-to-blame](https://thecorrespondent.com/751/weve-emitted-more-co2-in-the-past-30-years-than-in-all-of-history-these-three-reasons-are-to-blame.)
