---
title: "Analyzing Sustainability Indicators Amidst Climate Change"
subtitle: "A deeper analysis of CO₂"
author: 
  - "Eglah Belizor"
  - "Shamil Khusnullin"
  - "Mariia Koval"
abstract: "This page allows you to dive deeper into the exploration of CO₂ levels change happening over the last 30 years around the world. You can explore the CO₂ level changes on the map year by year. Furthermore, this page includes the clustering and a trained algorithm that predicts the CO₂ level change relative to 1990 based on the electricity production input."
resources:
  - Data/co2_map.rds
---

```{r}
#| label: setup
#| include: false

# Improve digit display and NA display for kable tables
options(scipen = 1, knitr.kable.NA = "")

# Load necessary packages
if (!require('pacman')) install.packages('pacman'); library('pacman')
pacman::p_load(viridis, tidyverse, ggplot2, lubridate, knitr, dplyr, purrr, 
               broom, sf, rnaturalearth, janitor, countrycode, stringr, 
               GGally, tidyr, kableExtra, shinylive, randomForest)

co2_map <- readRDS("Data/co2_map.rds")
data <- readRDS("Data/data.rds")
world <- ne_countries(scale = "medium", returnclass = "sf") # world map data
climate_long <- readRDS("Data/climate_long.rds")
```


## Interactive Map

The CO₂ levels change is a gradual process that happens over the year. To better understand the change, we’ve created an interactive map that lets you explore how CO₂ concentrations have shifted year by year, starting from the baseline year of 1991. By moving through the timeline, you can see which regions have experienced the most significant increases or decreases in emissions. 


<small>Use the slider below the map to select a specific year and observe the changes in CO₂ emissions compared to previous years. This allows you to closely examine how individual countries have increased or decreased their emissions over time. If you want full automated visualization: Click the play button to animate the map and watch the year-by-year evolution of global CO₂ emissions. </small>

<iframe src="https://shamchikkhus.shinyapps.io/MapApp/"
        width="100%" height="600px" frameborder="0">
</iframe>


## Classifying CO₂ changes
Next, we clustered the changes of CO₂ starting from 1991 up to 2015. Clustering on just one variable (percent change in CO₂ relative to 1990) makes the analysis very clear and direct and eliminates extra bias. By letting a clustering algorithm find its own group boundaries, we avoid guessing where “moderate” or “extreme” really begin. Instead, the data itself creates four natural clusters:

* Cluster 1: decrease or slight increase (<23.92% increase)


* Cluster 2: moderate increase (up to 119.4%)


* Cluster 3: high increase (up to 255%)


* Cluster 4: extreme increase (above 255%)


The plotted graph below should represent how these clusters looked like in 2015. You can compare this graph with the interactive map, choosing the year 2015, and see how clustering maps the graph.

```{r, warning=FALSE, fig.width=8, fig.height=4}
# Seed for Reproducibility
set.seed(631)

# Remove geometry and delete NA values
co2_cleaned <- co2_map |> 
  sf::st_drop_geometry(data_all) |> 
  drop_na() |> 
  select(`Country Code`, `Country Name`, `Year`, `co2conc`)
  

co2_numeric <- co2_cleaned |> 
  select(co2conc)

# Normalize the data
co2_scaled <- scale(co2_numeric) 

# Clustering
kmeans_result <- kmeans(co2_scaled, centers = 4, nstart = 25)


# Add cluster back to co2_cleaned df
co2_cleaned$Cluster <- as.factor(kmeans_result$cluster)

co2_final <- co2_cleaned |> 
  right_join(climate_long, by =  c("Country Name", "Country Code", "Year")) |> 
  drop_na() 

co2_final <- co2_final |> 
  mutate(Cluster = case_when(
    Cluster == "1" ~ "Extreme Increase",
    Cluster == "2" ~ "Decrease",
    Cluster == "3" ~ "Moderate Increase",
    Cluster == "4" ~ "High Increase"
  )) 

# Set Clusters as factor
co2_final <- co2_final |> 
  mutate(Cluster = as.factor(Cluster))



co2_final1 <- co2_final |> 
  mutate(sum = Renew_Elec + Oil_Elec + Coal_Elec + Gas_Elec)


set.seed(213)

# Train data on values from 1990 to 2014
train_data <- co2_final[co2_final$Year != 2015, ]


# Test data based on 2015
test_data <- co2_final[co2_final$Year == 2015,] 

# Build a Forest learning algorithm
# Use 4 different variables to predict a cluster
rf_model <- randomForest(Cluster ~ Renew_Cons + Renew_Elec + Oil_Elec + Gas_Elec
                         + Coal_Elec,
                         data = train_data, ntree = 100)

# Draw predictions for 2015 based on the model.
predictions <- predict(rf_model, test_data)

```
## Predicting CO₂ based on electricity production variables.

Furthermore, I build a machine learning model which predicts the cluster using 4 variables that we have discussed earlier:

-Renewable energy consumption (% of total final energy consumption)

-Renewable electricity output (% of total electricity output)

-Electricity production from oil sources (% of total)

-Electricity production from coal sources (% of total) 

You can observe the test on the actual data that the algorithm ran on, getting 78.6% accuracy.

```{r}

# Confusion Matrix
confusion_matrix <- table(Actual = test_data$Cluster, Predicted = predictions)
print(confusion_matrix)

# Calculate Accuracy
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
print(paste("Prediction Accuracy:", round(accuracy, 3)))


```

Furthermore, this page allows the policymakers to see how their proposed energy use system will impact the CO2 levels 

<iframe src="https://shamchikkhus.shinyapps.io/PredictionApp/"
        width="100%" height="600px" frameborder="0">
</iframe>
