---
title: "Analyzing Sustainability Indicators Amidst Climate Change"
subtitle: "A deeper analysis of CO₂"
author: 
  - "Eglah Belizor"
  - "Shamil Khusnullin"
  - "Mariia Koval"
abstract: "This page allows you to dive deeper into the exploration of CO₂ levels change happening over the last 30 years around the world. You can explore the CO₂ level changes on the map year by year. Furthermore, this page includes the clustering and a trained algorithm that predicts the CO₂ level change relative to 1990 based on the electricity production input."
resources:
  - Data/co2_map.rds
---

```{r}
#| label: setup
#| include: false

# Improve digit display and NA display for kable tables
options(scipen = 1, knitr.kable.NA = "")

# Load necessary packages
if (!require('pacman')) install.packages('pacman'); library('pacman')
pacman::p_load(viridis, tidyverse, ggplot2, lubridate, knitr, dplyr, purrr, 
               broom, sf, rnaturalearth, janitor, countrycode, stringr, 
               GGally, tidyr, kableExtra, shinylive, randomForest)

co2_map <- readRDS("Data/co2_map.rds")
data <- readRDS("Data/data.rds")
world <- ne_countries(scale = "medium", returnclass = "sf") # world map data
```


## Interactive Map

The CO₂ levels change is a gradual process that happens over the year. To better understand the change, we’ve created an interactive map that lets you explore how CO₂ concentrations have shifted year by year, starting from the baseline year of 1991. By moving through the timeline, you can see which regions have experienced the most significant increases or decreases in emissions. 


<small>Use the slider below the map to select a specific year and observe the changes in CO₂ emissions compared to previous years. This allows you to closely examine how individual countries have increased or decreased their emissions over time. If you want full automated visualization: Click the play button to animate the map and watch the year-by-year evolution of global CO₂ emissions. </small>

<iframe src="https://shamchikkhus.shinyapps.io/MapApp/"
        width="100%" height="600px" frameborder="0">
</iframe>



## Map by clustering

We have clustered
Continuing on the topic of clusters, I've decided the cluster the 2015 CO2 data and plot the graph based on 4 levels of CO₂ change. The new graph gives a clear persepctive on how countries' CO₂ levels have changed. For example, the decrease in CO₂ is prelevant in the developed countries, while the developing countries have experienced moderate to extreme CO2 levels increase. This increase can be atrributed to a very rapid industralization happening in those counties.

```{r, warning=FALSE, fig.width=8, fig.height=4}
# Set seed for reproducibility
set.seed(631)
# Filter for year 2022
co2_2015 <- st_as_sf(co2_map) |> 
  filter(Year == 2015) |> 
   drop_na() 

co2_2015_values <- co2_2015 |> 
  sf::st_drop_geometry(data_all) |> 
  select(co2conc)


# Standardize the data
co2_2015_scaled <- scale(co2_2015_values) 

# Cluster the data
kmeans_result <- kmeans(co2_2015_scaled, centers = 4, nstart = 25)

# Insert clusters back a variable
co2_2015$Cluster <- as.factor(kmeans_result$cluster)


co2_2015 <- co2_2015 |> 
 mutate(Cluster = case_when(
 Cluster == "1" ~ "Moderate Increase",
    Cluster == "2" ~ "Extreme Increase",
    Cluster == "3" ~ "Decrease/Slight Increase",
    Cluster == "4" ~ "High Increase",
))

# Custom color mapping
cluster_colors <- c(
  "Extreme Increase"  = "#d73027",  
  "High Increase"     = "#fc8d59",
  "Moderate Increase" = "#fee08b",  
  "Decrease/Slight Increase" = "#1a9850" 
)

# Plot
ggplot(co2_2015) +
  geom_sf(aes(fill = Cluster), color = "grey40", linewidth = 0.2) +
  coord_sf(crs = "+proj=robin", expand = FALSE) +
  scale_fill_manual(values = cluster_colors)+
  theme_void(base_size = 14) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 8)
  ) +
  labs(title = "2015 CO2 emission Change relative to 1990")

```

```{r}
# Seed for Reproducibility
set.seed(631)

# Download the data

co2_map <- readRDS("raw-data/co2_map.rds")
climate1 <- read.csv("raw-data/climate-change.csv", skip = 3)
climate1 <- climate1 |>
  row_to_names(row_number = 1)|>
  select(-"Indicator Code") 

climate_long <- climate1 |> 
  # 1) Filter for the four indicators in the image
  filter(
    `Indicator Name` %in% c(
      "Renewable energy consumption (% of total final energy consumption)",
      "Renewable electricity output (% of total electricity output)",
      "Electricity production from oil sources (% of total)",
      "Electricity production from coal sources (% of total)"
    )
  ) |> 
  # 2) Keep only the essential columns
  select(
    "Country Code",
    "Country Name",
    "Indicator Name",
    matches("^(19|20)\\d{2}$")
  ) |> 
  
  mutate(across(matches("^(19|20)\\d{2}$"), as.numeric)) |> 
  # 5) Pivot longer to get Year and Value
  pivot_longer(
    cols      = matches("^(19|20)\\d{2}$"),
    names_to  = "Year",
    values_to = "Value"
  ) |> 
  # 6) Pivot wider to separate each indicator into its own column
  pivot_wider(
    names_from  = `Indicator Name`,
    values_from = Value
  ) |> 
  drop_na() |> 
  mutate(Year = as.numeric(Year))

names(climate_long) <- c("Country Code", "Country Name", "Year",
                         "Renew_Cons", "Renew_Elec",   "Oil_Elec",
                         "Coal_Elec")



# Remove geometry and delete NA values
co2_cleaned <- co2_map |> 
  sf::st_drop_geometry(data_all) |> 
  drop_na() |> 
  select(`Country Code`, `Country Name`, `Year`, `co2conc`)
  
# Leave only numeric variable for clustering
co2_numeric <- co2_cleaned |> 
  select(co2conc)

# Scale the data
co2_scaled <- scale(co2_numeric) 

# Clustering
kmeans_result <- kmeans(co2_scaled, centers = 4, nstart = 25)

co2_cleaned$Cluster <- as.factor(kmeans_result$cluster)

co2_final <- co2_cleaned |> 
  right_join(climate_long, by =  c("Country Name", "Country Code", "Year")) |> 
  drop_na() 

co2_final <- co2_final |> 
  mutate(Cluster = case_when(
    Cluster == "1" ~ "Extreme Increase",
    Cluster == "2" ~ "Decrease",
    Cluster == "3" ~ "Moderate Increase",
    Cluster == "4" ~ "High Increase"
  )) 

# Set Clusters as factor
co2_final <- co2_final |> 
  mutate(Cluster = as.factor(Cluster))





# Choose a training sample of 70% 
sample_index <- sample(1:nrow(co2_final), 0.7 * nrow(co2_final))

train_data <- co2_final[sample_index, ]
#Choose a testing sample of the remaining 30%
test_data <- co2_final[-sample_index, ]

# Build a Forest learning algorithm
rf_model <- randomForest(Cluster ~ Renew_Cons + Renew_Elec + Oil_Elec 
                         + Coal_Elec,
                         data = train_data, ntree = 100)

# Draw predictions based on the previous table
predictions <- predict(rf_model, test_data)

```
## Machine Learning

Furthermore, I build a machine learning model which predicts the cluster using 4 variables that we have discussed earlier:

-Renewable energy consumption (% of total final energy consumption)

-Renewable electricity output (% of total electricity output)

-Electricity production from oil sources (% of total)

-Electricity production from coal sources (% of total) 

You can observe the test on the actual data that the algorithm ran on, getting 78.6% accuracy.

```{r}

# Confusion Matrix
confusion_matrix <- table(Actual = test_data$Cluster, Predicted = predictions)
print(confusion_matrix)

# Calculate Accuracy
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
print(paste("Prediction Accuracy:", round(accuracy, 3)))


```

Furthermore, this page allows the policymakers to see how their proposed energy use system will impact the CO2 levels 

<iframe src="https://shamchikkhus.shinyapps.io/PredictionApp/"
        width="100%" height="600px" frameborder="0">
</iframe>
