---
title: "Analyzing Sustainability Indicators Amidst Climate Change"
author: 
  - "Eglah Belizor"
  - "Shamil Khusnullin"
  - "Mariia Koval"
resources:
  - Data/co2_map.rds
---

```{r}
#| label: setup
#| include: false

# Improve digit display and NA display for kable tables
options(scipen = 1, knitr.kable.NA = "")

# Load necessary packages
if (!require('pacman')) install.packages('pacman'); library('pacman')
pacman::p_load(viridis, tidyverse, ggplot2, lubridate, knitr, dplyr, purrr, 
               broom, sf, rnaturalearth, janitor, countrycode, stringr, 
               GGally, tidyr, kableExtra, shinylive, randomForest)

co2_map <- readRDS("Data/co2_map.rds")
data <- readRDS("Data/data.rds")
```

# Clustering by sustainability levels

We clustered the data between 1990 and 2015 to determine country's sustainability by 4 clusters and using the following variables:

* Renewable energy consumption (% of total final energy consumption)

* Renewable electricity output (% of total electricity output)

* Electricity production from oil sources (% of total)

* Electricity production from coal sources (% of total) 


(See [Appendix](#appendix) to see why we've chosen k=4 for clustering).

```{r, fig.show='hide'}
#| label: kmeans-clustering

climate1 <- read.csv("raw-data/climate-change.csv", skip = 3)
climate1 <- climate1 |>
  row_to_names(row_number = 1)|>
  select(-"Indicator Code") 

# vector for selected variables of interest
sustain_indicators <- c(
  "Renewable energy consumption (% of total final energy consumption)",
  "Renewable electricity output (% of total electricity output)",
  "Electricity production from oil sources (% of total)",
  "Electricity production from coal sources (% of total)")

# setting up first period data frame (year 1990)
sustainability1990 <- climate1 |>
    filter(`Indicator Name` %in% sustain_indicators) |>
    select(`Country Name`, `Country Code`, `Indicator Name`, `1990`) |>
    pivot_wider(names_from = `Indicator Name`, 
                values_from = `1990`) |>
    drop_na()

# setting up second period data frame (year 2015)
sustainability2015 <- climate1 |>
    filter(`Indicator Name` %in% sustain_indicators) |>
    select(`Country Name`, `Country Code`, `Indicator Name`, `2015`) |>
    pivot_wider(names_from = `Indicator Name`, 
                values_from = `2015`) |>
    drop_na()

# standardize the indicator values for 1990
sustain_1990_std <- sustainability1990 |>
  mutate(across(where(is.numeric)
                , ~ (.x - mean(.x)) / sd(.x)
                , .names = "{.col}_z")) |>
  select(ends_with("_z"))

# standardize the indicator values for 2015
sustain_2015_std <- sustainability2015 |>
  mutate(across(where(is.numeric)
                , ~ (.x - mean(.x)) / sd(.x)
                , .names = "{.col}_z")) |>
  select(ends_with("_z"))

# run k-means clustering
set.seed(231)

# first, elbow plots for each year's standardized data
elbow_1990 <- tibble(k = 1:10) |>
  mutate(
    kmeans_results = map(k, ~kmeans(sustain_1990_std, .x, nstart = 20)),
    glanced = map(kmeans_results, glance)
  ) |>
  unnest(cols = c(glanced))
elbow_2015 <- tibble(k = 1:10) |>
  mutate(
    kmeans_results = map(k, ~kmeans(sustain_2015_std, .x, nstart = 20)),
    glanced = map(kmeans_results, glance)
  ) |>
  unnest(cols = c(glanced))

ggplot(elbow_1990, aes(x = k, y = tot.withinss)) +
  geom_point() + 
  geom_line() +
  scale_x_continuous(breaks = 1:10) +
  labs(title = "Elbow Plot for 1990 Sustainability Data",
       x = "Number of Clusters (k)", 
       y = "Total Within-Cluster Sum of Squares") +
  theme_minimal()
ggplot(elbow_2015, aes(x = k, y = tot.withinss)) +
  geom_point() + 
  geom_line() +
  scale_x_continuous(breaks = 1:10) +
  labs(title = "Elbow Plot for 2015 Sustainability Data",
       x = "Number of Clusters (k)", 
       y = "Total Within-Cluster Sum of Squares") +
  theme_minimal()

# based on the elbow plot, k = 4 (output was hidden from this page but image of the plots can be found in Appendix)
sustain_1990_kmeans4 <- sustain_1990_std |>
  kmeans(centers = 4, nstart = 20)
sustain_2015_kmeans4 <- sustain_2015_std |>
  kmeans(centers = 4, nstart = 20)

sustain_1990_c4 <- augment(sustain_1990_kmeans4, sustainability1990) |>
  rename(cluster_1990 = .cluster)
sustain_2015_c4 <- augment(sustain_2015_kmeans4, sustainability2015) |>
  rename(cluster_2015 = .cluster)

# comparing clusters
clusters_1990 <- sustain_1990_c4 |>
  select(`Country Name`, cluster_1990)
clusters_2015 <- sustain_2015_c4 |>
  select(`Country Name`, cluster_2015)
```

After the process of standardizing indicators and k-means clustering for the years 1990 and 2015 was complete, the table we created allowed us to take a glimpse at the changes in cluster group assignments for the first 10 countries and regions in this newly formed dataset. At first glance, you can see the region of 'Africa Western and Central' changed groups from 3 to 2, while the country of Argentina moved from 4 to 3.

```{r}
cluster_comparison <- clusters_1990 |>
  inner_join(clusters_2015, by = "Country Name") |>
  rename(`1990 Cluster` = `cluster_1990`,
         `2015 Cluster` = `cluster_2015`) |>
  mutate(Moved = `1990 Cluster` != `2015 Cluster`) #Count how many countries moved clusters


# kable table
cluster_comparison |>
  kbl(booktabs = TRUE, align = "l") |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = TRUE,
    position = "center",
    font_size = 12) |> 
  scroll_box(width = "500px", height = "200px")
```

## Scatterplot matrices of sustainability indicators by cluster

Next, the goal was to visualize the distribution of values for each cluster group in order to distinguish each group's unique characteristics using a scatterplot matrix. There are two matrices to represent the cluster groups for two different periods. By knowing each cluster's special characteristics, we are able to identify which groups are most sustainable and the least sustainable.

::: {.panel-tabset}

### 1990

```{r}
# Scatterplot matrix for standardized 1990 cluster data
sustain_1990_c4 |>
  rename (
      renew_energy_consume =
        "Renewable energy consumption (% of total final energy consumption)",
      renew_electric_output = 
        "Renewable electricity output (% of total electricity output)",
      electric_prod_oil = 
        "Electricity production from oil sources (% of total)", 
      electric_prod_coal = 
        "Electricity production from coal sources (% of total)"
    ) |>
  select(renew_energy_consume, renew_electric_output, electric_prod_oil, 
         electric_prod_coal, cluster_1990) |>
  ggpairs(aes(color = cluster_1990),
          upper = list(continuous = "blank"),
          progress = FALSE) +
  theme(text = element_text(size = 8))
```

### 2015

```{r}
# Scatterplot matrix for standardized 2015 cluster data
sustain_2015_c4 |>
  rename (
      renew_energy_consume =
        "Renewable energy consumption (% of total final energy consumption)",
      renew_electric_output = 
        "Renewable electricity output (% of total electricity output)",
      electric_prod_oil = 
        "Electricity production from oil sources (% of total)", 
      electric_prod_coal =
        "Electricity production from coal sources (% of total)"
    ) |>
  select(renew_energy_consume, renew_electric_output, electric_prod_oil, 
         electric_prod_coal, cluster_2015) |>
  ggpairs(aes(color = cluster_2015),
          upper = list(continuous = "blank"),
          progress = FALSE) +
  theme(text = element_text(size = 8))
```

:::

In 1990:

- **Cluster 1 (red)**  
  - Relatively lower values across all four indicators. Although two outliers with higher oil production.
  - Moderate sustainability: Not a stand out in fossil fuel or renewable usage.

- **Cluster 2 (green)**  
  - Characterized by high oil-based electricity production, but very low renewable energy consumption and renewable electricity output.
  - Low- to moderate sustainability: Higher oil-based electric production than the other clusters.

- **Cluster 3 (blue)**  
  - Higher levels of renewable energy consumption and electricity output, with low oil and coal electricity production.
  - Most sustainable: Likely represents countries investing in renewables with limited reliance on fossil fuels.

- **Cluster 4 (purple)**  
  - All countries have relatively high coal-based electricity production, while a few of them have moderate renewable energy consumption. 
  - Least sustainable: Higher fossil fuel usage (coal) than the others.

**Ranking from least to most sustainable groups in 1990**: 4, 2, 1, 3

In 2015:

- **Cluster 1 (red)**  
  - Although relatively higher renewable energy consumption, there is very high oil-based production.
  - Low- to moderate sustainable group. 

- **Cluster 2 (green)**  
  - Now the most sustainable. Relatively high renewable energy consumption and output.

- **Cluster 3 (blue)**  
  - Lower values across all four indicators.  
  - Now the moderately sustainable group.

- **Cluster 4 (purple)**  
  - Higher coal production than the other clusters. Moderate renewable energy and output.  
  - Least sustainable due to high coal-based production.

**Ranking from least to most sustainable groups in 2015**: 4, 1, 3, 2

## Sustainability cluster map

Holistic Summary of cluster data based on the world map: From the available data, there were more countries and regions in the 'least sustainable' group (34) than countries in the 'most sustainable' group (19) in 1990. The most countries were in the 'moderately sustainable' group (40). By 2015, more countries entered the most sustainable group (43), while the least sustainable countries were the smallest group (14). See the Appendix for more information on the exact amount of countries in each group for both periods.
Interestingly, there is much regional and economic diversity in countries who share the same group. In 2015, many Sub-Saharan African countries were most sustainable along with countries like Canada, Brazil, and Scandinavia (Denmark, Norway, Sweden, Finland).

::: {.panel-tabset}

### 1990 Map
```{r}
library(plotly)

# Prepare your sustainability + cluster data (already renamed):
data_1990 <- sustainability1990 |>
  left_join(clusters_1990, by = c("Country Name")) |>
  rename(
    `Renewable Energy (%)` = `Renewable energy consumption (% of total final energy consumption)`,
    `Renewable Electricity (%)` = `Renewable electricity output (% of total electricity output)`,
    `Oil Electricity (%)` = `Electricity production from oil sources (% of total)`,
    `Coal Electricity (%)` = `Electricity production from coal sources (% of total)`
  ) |>
  mutate(
    cluster_label = factor(cluster_1990,
                           levels = c(4, 2, 1, 3),
                           labels = c("Least Sustainable", "Low-Moderate", "Moderate", "Most Sustainable")),
    hover_text = paste0(
      "<b>", `Country Name`, "</b><br>",
      "Cluster: ", cluster_label, "<br>",
      "Renewable Energy: ", round(`Renewable Energy (%)`, 1), "%<br>",
      "Renewable Electricity: ", round(`Renewable Electricity (%)`, 1), "%<br>",
      "Oil Electricity: ", round(`Oil Electricity (%)`, 1), "%<br>",
      "Coal Electricity: ", round(`Coal Electricity (%)`, 1), "%"
    )
  )

# Map using ISO country codes
plot_1990 <- plot_geo(data_1990) |>
  add_trace(
    locations = ~`Country Code`,
    locationmode = "ISO-3",
    z = ~as.numeric(factor(cluster_1990, levels = c(4, 2, 1, 3))),
    colors = c("1" = "#FF0000", "4" = "#FF9900", "3" = "#FFFF00", "2" = "#33CC33"),
    text = ~hover_text,
    hoverinfo = "text",
    showscale = FALSE
  ) |>
  layout(
    title = "Sustainability Clusters and Indicators in 1990",
    geo = list(showframe = TRUE, showcoastlines = TRUE)
  )
plot_1990

```

### 2015 Map
```{r}
# Prep 2015 data
data_2015 <- sustainability2015 |>
  left_join(clusters_2015, by = c("Country Name")) |>
  rename(
    `Renewable Energy (%)` = `Renewable energy consumption (% of total final energy consumption)`,
    `Renewable Electricity (%)` = `Renewable electricity output (% of total electricity output)`,
    `Oil Electricity (%)` = `Electricity production from oil sources (% of total)`,
    `Coal Electricity (%)` = `Electricity production from coal sources (% of total)`
  ) |>
  mutate(
    cluster_label = factor(cluster_2015,
                           levels = c(4, 1, 3, 2),
                           labels = c("Least Sustainable", "Low-Moderate", "Moderate", "Most Sustainable")),
    hover_text = paste0(
      "<b>", `Country Name`, "</b><br>",
      "Cluster: ", cluster_label, "<br>",
      "Renewable Energy: ", round(`Renewable Energy (%)`, 1), "%<br>",
      "Renewable Electricity: ", round(`Renewable Electricity (%)`, 1), "%<br>",
      "Oil Electricity: ", round(`Oil Electricity (%)`, 1), "%<br>",
      "Coal Electricity: ", round(`Coal Electricity (%)`, 1), "%"
    )
  )

# Map (2015)
plot_2015 <- plot_geo(data_2015) |>
  add_trace(
    locations = ~`Country Code`,
    locationmode = "ISO-3",
    z = ~as.numeric(factor(cluster_2015, levels = c(4, 1, 3, 2))),  # force correct mapping order
    colors = c("#FF0000", "#FF9900", "#FFFF00", "#33CC33"),        # red, orange, yellow, green
    text = ~hover_text,
    hoverinfo = "text",
    showscale = FALSE
  ) |>
  layout(
    title = "Sustainability Clusters and Indicators in 2015",
    geo = list(showframe = TRUE, showcoastlines = TRUE)
  )

plot_2015

```

:::

## References
* https://cran.r-project.org/web/packages/rnaturalearth/vignettes/rnaturalearth.html
* https://html-color-codes.com/
