---
title: "Analyzing Sustainability Indicators Amidst Climate Change"
author: 
  - "Eglah Belizor"
  - "Shamil Khusnullin"
  - "Mariia Koval"
resources:
  - Data/co2_map.rds
---

```{r}
#| label: setup
#| include: false

# Improve digit display and NA display for kable tables
options(scipen = 1, knitr.kable.NA = "")

# Load necessary packages
if (!require('pacman')) install.packages('pacman'); library('pacman')
pacman::p_load(viridis, tidyverse, ggplot2, lubridate, knitr, dplyr, purrr, 
               broom, sf, rnaturalearth, janitor, countrycode, stringr, 
               GGally, tidyr, kableExtra, shinylive, randomForest)

co2_map <- readRDS("Data/co2_map.rds")
data <- readRDS("Data/data.rds")
world <- ne_countries(scale = "medium", returnclass = "sf") # world map data
```

# Clustering by sustainability levels

We clustered the data between 1990 and 2015 to determine country's sustainability by 4 clusters and using the following variables:

-Renewable energy consumption (% of total final energy consumption)

-Renewable electricity output (% of total electricity output)

-Electricity production from oil sources (% of total)

-Electricity production from coal sources (% of total) 


(See appendix to see why we've chose k=4 for clustering).

```{r}
#| label: kmeans-clustering

climate1 <- read.csv("raw-data/climate-change.csv", skip = 3)
climate1 <- climate1 |>
  row_to_names(row_number = 1)|>
  select(-"Indicator Code") 

# vector for selected variables of interest
sustain_indicators <- c(
  "Renewable energy consumption (% of total final energy consumption)",
  "Renewable electricity output (% of total electricity output)",
  "Electricity production from oil sources (% of total)",
  "Electricity production from coal sources (% of total)")

# setting up first period data frame (year 1990)
sustainability1990 <- climate1 |>
    filter(`Indicator Name` %in% sustain_indicators) |>
    select(`Country Name`, `Country Code`, `Indicator Name`, `1990`) |>
    pivot_wider(names_from = `Indicator Name`, 
                values_from = `1990`) |>
    drop_na()

# setting up second period data frame (year 2015)
sustainability2015 <- climate1 |>
    filter(`Indicator Name` %in% sustain_indicators) |>
    select(`Country Name`, `Country Code`, `Indicator Name`, `2015`) |>
    pivot_wider(names_from = `Indicator Name`, 
                values_from = `2015`) |>
    drop_na()

# standardize the indicator values for 1990
sustain_1990_std <- sustainability1990 |>
  mutate(across(where(is.numeric)
                , ~ (.x - mean(.x)) / sd(.x)
                , .names = "{.col}_z")) |>
  select(ends_with("_z"))

# standardize the indicator values for 2015
sustain_2015_std <- sustainability2015 |>
  mutate(across(where(is.numeric)
                , ~ (.x - mean(.x)) / sd(.x)
                , .names = "{.col}_z")) |>
  select(ends_with("_z"))

# run k-means clustering
set.seed(231)

# based on the elbow plot, k = 4
sustain_1990_kmeans4 <- sustain_1990_std |>
  kmeans(centers = 4, nstart = 20)
sustain_2015_kmeans4 <- sustain_2015_std |>
  kmeans(centers = 4, nstart = 20)

sustain_1990_c4 <- augment(sustain_1990_kmeans4, sustainability1990) |>
  rename(cluster_1990 = .cluster)
sustain_2015_c4 <- augment(sustain_2015_kmeans4, sustainability2015) |>
  rename(cluster_2015 = .cluster)

# comparing clusters
clusters_1990 <- sustain_1990_c4 |>
  select(`Country Name`, cluster_1990)
clusters_2015 <- sustain_2015_c4 |>
  select(`Country Name`, cluster_2015)

world_clusters <- world |>
  left_join(clusters_1990, by = c("name" = "Country Name")) |>
  left_join(clusters_2015, by = c("name" = "Country Name"))

```

After the process of standardizing indicators and k-means clustering for the years 1990 and 2015 was complete, the table we created allowed us to take a glimpse at the changes in cluster group assignments for the first 10 countries and regions in this newly formed dataset. At first glance, you can see the region of 'Africa Eastern and Southern' changed groups from 3 to 4, while the country of Argentina moved from 4 to 3.

```{r}
cluster_comparison <- clusters_1990 |>
  inner_join(clusters_2015, by = "Country Name") |>
  rename(`1990 Cluster` = `cluster_1990`,
         `2015 Cluster` = `cluster_2015`)

# kable table
cluster_comparison |>
  kbl(booktabs = TRUE, align = "l") |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = TRUE,
    position = "center",
    font_size = 12) |> 
  scroll_box(width = "500px", height = "200px")
```

Next, the goal was to visualize the distribution of values for each cluster group in order to distinguish each group's unique characteristics using a scatterplot matrix. There are two matrices to represent the cluster groups for two different periods. By knowing each cluster's special characteristics, we are able to identify which groups are most sustainable and the least sustainable.

```{r}
# visualization choice: scatterplot matrix
# for standardized 1990 cluster data
sustain_1990_c4 |>
  rename (
      renew_energy_consume =
        "Renewable energy consumption (% of total final energy consumption)",
      renew_electric_output = 
        "Renewable electricity output (% of total electricity output)",
      electric_prod_oil = 
        "Electricity production from oil sources (% of total)", 
      electric_prod_coal = 
        "Electricity production from coal sources (% of total)"
    ) |>
  select(`renew_energy_consume`, `renew_electric_output`, `electric_prod_oil`, 
         `electric_prod_coal`, `cluster_1990`) |>
  ggpairs(aes(color = cluster_1990)
        , upper = list(continuous = "blank")
        , progress = FALSE) +
  theme(text = element_text(size = 8))

# for standardized 2015 cluster data
sustain_2015_c4 |>
  rename (
      renew_energy_consume =
        "Renewable energy consumption (% of total final energy consumption)",
      renew_electric_output = 
        "Renewable electricity output (% of total electricity output)",
      electric_prod_oil = 
        "Electricity production from oil sources (% of total)", 
      electric_prod_coal =
        "Electricity production from coal sources (% of total)"
    ) |>
  select(`renew_energy_consume`, `renew_electric_output`, `electric_prod_oil`, 
         `electric_prod_coal`, `cluster_2015`) |>
  ggpairs(aes(color = cluster_2015)
        , upper = list(continuous = "blank")
        , progress = FALSE) +
  theme(text = element_text(size = 8))
```

In 1990:

- **Cluster 1 (red)**  
  - Characterized by high oil-based electricity production, but very low renewable energy consumption and renewable electricity output.  
  - Least sustainable: It likely represents countries heavily reliant on fossil fuels with minimal use of renewables.

- **Cluster 2 (green)**  
  - Higher levels of renewable energy consumption and electricity output, with low oil and coal electricity production.  
  - Most sustainable: Likely represents countries investing in renewables with limited reliance on fossil fuels.

- **Cluster 3 (blue)**  
  - Relatively high coal-based electricity production, with moderate renewable energy consumption.  
  - Second most sustainable: Notably high coal usage, but more renewables than Cluster 1.

- **Cluster 4 (purple)**  
  - Low to moderate values across all four indicators. Moderate coal usage but low in renewable energy usage.  
  - Third most sustainable: Not a standout in fossil fuel or renewable use.

**Ranking from most to least sustainable**: 2, 3, 4, 1

In 2015:

- **Cluster 1 (red)**  
  - Remains the least sustainable, although with some improvement in renewable energy consumption.

- **Cluster 2 (green)**  
  - Remains the most sustainable. Clear leader in sustainability metrics.

- **Cluster 3 (blue)**  
  - Reduced reliance on coal production, slight increase in renewables.  
  - Still in second place in sustainability ranking.

- **Cluster 4 (purple)**  
  - Increased coal production, surpassing Cluster 3. Moderate renewable energy, low oil usage.  
  - Still in third place in sustainability ranking.
  
### Sustainability cluster map

```{r}
# 1990 cluster map
ggplot(world_clusters) +
  geom_sf(aes(fill = as.factor(cluster_1990)), color = "black", size = 0.1) +
  scale_fill_brewer(na.value = "grey90", name = "Cluster 1990") +
  theme_minimal() +
  labs(title = "Country Clusters by Sustainability Levels in 1990")

# 2015 cluster map
ggplot(world_clusters) +
  geom_sf(aes(fill = as.factor(cluster_2015)), color = "black", size = 0.1) +
  scale_fill_brewer(na.value = "grey90", name = "Cluster 2015") +
  theme_minimal() +
  labs(title = "Country Clusters by Sustainability Levels in 2015")

```


The elbow plots below were used to determine the optimal number of cluster groups (k) for countries based on their sustainability levels. Not all seven sustainability indicators were considered in the k-means clustering model because the goal was to prioritize the sustainability indicators that signal robust economic activity and is thus prone to annual fluctuations, such as renewable energy consumption, renewable electricity output, as well as electricity production from oil and coal sources. We determined that 4 cluster groups was the optimal amount to proceed with our analysis.

``` {r}
#| label: elbow-plots

# first, elbow plots for each year's standardized data
elbow_1990 <- tibble(k = 1:10) |>
  mutate(
    kmeans_results = map(k, ~kmeans(sustain_1990_std, .x, nstart = 20)),
    glanced = map(kmeans_results, glance)
  ) |>
  unnest(cols = c(glanced))
elbow_2015 <- tibble(k = 1:10) |>
  mutate(
    kmeans_results = map(k, ~kmeans(sustain_2015_std, .x, nstart = 20)),
    glanced = map(kmeans_results, glance)
  ) |>
  unnest(cols = c(glanced))

# Plot an elbow graph to identify the best k value
ggplot(elbow_1990, aes(x = k, y = tot.withinss)) +
  geom_point() + 
  geom_line() +
  scale_x_continuous(breaks = 1:10) +
  labs(title = "Elbow Plot for 1990 Sustainability Data",
       x = "Number of Clusters (k)", 
       y = "Total Within-Cluster Sum of Squares") +
  theme_minimal()
ggplot(elbow_2015, aes(x = k, y = tot.withinss)) +
  geom_point() + 
  geom_line() +
  scale_x_continuous(breaks = 1:10) +
  labs(title = "Elbow Plot for 2015 Sustainability Data",
       x = "Number of Clusters (k)", 
       y = "Total Within-Cluster Sum of Squares") +
  theme_minimal()
```
