---
title: "Analyzing Sustainability Indicators Amidst Climate Change"
author: 
  - "Eglah Belizor"
  - "Shamil Khusnullin"
  - "Mariia Koval"
---

```{r}
#| label: setup
#| include: false

# Improve digit display and NA display for kable tables
options(scipen = 1, knitr.kable.NA = "")

# Load necessary packages
if (!require('pacman')) install.packages('pacman'); library('pacman')
pacman::p_load(viridis, tidyverse, ggplot2, lubridate, knitr, dplyr, purrr, 
               broom, sf, rnaturalearth, janitor, countrycode, stringr, 
               GGally, tidyr, kableExtra, shinylive, randomForest)

```

# Clustering by sustainability levels

To better understand global trends in energy sustainability over time, we applied k-means clustering to a World Bank dataset spanning from 1990 to 2015. Our objective was to classify countries and regions into one of four groups based on their energy production and consumption profiles. The analysis focused on the following four key energy indicators:

* `Renewable energy consumption (% of total final energy consumption)`

* `Renewable electricity output (% of total electricity output)`

* `Electricity production from oil sources (% of total)`

* `Electricity production from coal sources (% of total)`


(See [Appendix](#appendix) to see why we've chosen k=4 for clustering).

```{r, fig.show='hide'}
#| label: kmeans-clustering

climate1 <- read.csv("raw-data/climate-change.csv", skip = 3)
climate1 <- climate1 |>
  row_to_names(row_number = 1)|>
  select(-"Indicator Code") 

# vector for selected variables of interest
sustain_indicators <- c(
  "Renewable energy consumption (% of total final energy consumption)",
  "Renewable electricity output (% of total electricity output)",
  "Electricity production from oil sources (% of total)",
  "Electricity production from coal sources (% of total)")

# setting up first period data frame (year 1990)
sustainability1990 <- climate1 |>
    filter(`Indicator Name` %in% sustain_indicators) |>
    select(`Country Name`, `Country Code`, `Indicator Name`, `1990`) |>
    pivot_wider(names_from = `Indicator Name`, 
                values_from = `1990`) |>
    drop_na()

# setting up second period data frame (year 2015)
sustainability2015 <- climate1 |>
    filter(`Indicator Name` %in% sustain_indicators) |>
    select(`Country Name`, `Country Code`, `Indicator Name`, `2015`) |>
    pivot_wider(names_from = `Indicator Name`, 
                values_from = `2015`) |>
    drop_na()

# standardize the indicator values for 1990
sustain_1990_std <- sustainability1990 |>
  mutate(across(where(is.numeric)
                , ~ (.x - mean(.x)) / sd(.x)
                , .names = "{.col}_z")) |>
  select(ends_with("_z"))

# standardize the indicator values for 2015
sustain_2015_std <- sustainability2015 |>
  mutate(across(where(is.numeric)
                , ~ (.x - mean(.x)) / sd(.x)
                , .names = "{.col}_z")) |>
  select(ends_with("_z"))

# run k-means clustering
set.seed(231)

# first, elbow plots for each year's standardized data
elbow_1990 <- tibble(k = 1:10) |>
  mutate(
    kmeans_results = map(k, ~kmeans(sustain_1990_std, .x, nstart = 20)),
    glanced = map(kmeans_results, glance)
  ) |>
  unnest(cols = c(glanced))
elbow_2015 <- tibble(k = 1:10) |>
  mutate(
    kmeans_results = map(k, ~kmeans(sustain_2015_std, .x, nstart = 20)),
    glanced = map(kmeans_results, glance)
  ) |>
  unnest(cols = c(glanced))

ggplot(elbow_1990, aes(x = k, y = tot.withinss)) +
  geom_point() + 
  geom_line() +
  scale_x_continuous(breaks = 1:10) +
  labs(title = "Elbow Plot for 1990 Sustainability Data",
       x = "Number of Clusters (k)", 
       y = "Total Within-Cluster Sum of Squares") +
  theme_minimal()
ggplot(elbow_2015, aes(x = k, y = tot.withinss)) +
  geom_point() + 
  geom_line() +
  scale_x_continuous(breaks = 1:10) +
  labs(title = "Elbow Plot for 2015 Sustainability Data",
       x = "Number of Clusters (k)", 
       y = "Total Within-Cluster Sum of Squares") +
  theme_minimal()

# based on the elbow plot, k = 4 (output was hidden from this page but image of the plots can be found in Appendix)
sustain_1990_kmeans4 <- sustain_1990_std |>
  kmeans(centers = 4, nstart = 20)
sustain_2015_kmeans4 <- sustain_2015_std |>
  kmeans(centers = 4, nstart = 20)

sustain_1990_c4 <- augment(sustain_1990_kmeans4, sustainability1990) |>
  rename(cluster_1990 = .cluster)
sustain_2015_c4 <- augment(sustain_2015_kmeans4, sustainability2015) |>
  rename(cluster_2015 = .cluster)

# comparing clusters
clusters_1990 <- sustain_1990_c4 |>
  select(`Country Name`, cluster_1990)
clusters_2015 <- sustain_2015_c4 |>
  select(`Country Name`, cluster_2015)
```

By standardizing these indicators and conducting clustering separately for each year, we were able to group countries with similar energy characteristics and track how these groupings evolved over time. A table comparing cluster assignments for all countries and regions with publicly available data revealed some notable changes. For example, Africa Eastern and Southern shifted from cluster 3 to cluster 4, and Argentina moved from cluster 4 to cluster 3 between 1990 and 2015. However, a change in cluster assignment does not always reflect real changes in sustainability. Australia, for instance, moved from cluster 3 to cluster 4, yet both clusters represented the least sustainable group in their respective years.

Since k-means clustering is conducted independently for each time period, the composition and characteristics of each cluster may vary. Therefore, it is necessary to evaluate and rank the clusters separately within each year to draw accurate and meaningful conclusions about sustainability progress over time.

```{r}
cluster_comparison <- clusters_1990 |>
  inner_join(clusters_2015, by = "Country Name") |>
  rename(`1990 Cluster` = `cluster_1990`,
         `2015 Cluster` = `cluster_2015`) |>
  mutate(Moved = `1990 Cluster` != `2015 Cluster`) #Count how many countries moved clusters

# kable table
cluster_comparison |>
  kbl(booktabs = TRUE, align = "l") |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = TRUE,
    position = "center",
    font_size = 12) |> 
  scroll_box(width = "500px", height = "200px")
```

## Summary statistics for sustainability cluster groups

To better understand the defining characteristics of each cluster, we next visualized the distribution of values for the four energy indicators using scatterplot matrices for 1990 and 2015. These visualizations allowed us to differentiate clusters based on their energy profiles and to assign relative sustainability rankings.

::: {.panel-tabset}

### 1990
```{r}
# Scatterplot matrix for standardized 1990 cluster data
sustain_1990_c4 |>
  rename (
      renew_energy_consume =
        "Renewable energy consumption (% of total final energy consumption)",
      renew_electric_output = 
        "Renewable electricity output (% of total electricity output)",
      electric_prod_oil = 
        "Electricity production from oil sources (% of total)", 
      electric_prod_coal = 
        "Electricity production from coal sources (% of total)"
    ) |>
  select(renew_energy_consume, renew_electric_output, electric_prod_oil, 
         electric_prod_coal, cluster_1990) |>
  ggpairs(aes(color = cluster_1990),
          upper = list(continuous = "blank"),
          progress = FALSE) +
  theme(text = element_text(size = 8))
```

### 2015
```{r}
# Scatterplot matrix for standardized 2015 cluster data
sustain_2015_c4 |>
  rename (
      renew_energy_consume =
        "Renewable energy consumption (% of total final energy consumption)",
      renew_electric_output = 
        "Renewable electricity output (% of total electricity output)",
      electric_prod_oil = 
        "Electricity production from oil sources (% of total)", 
      electric_prod_coal =
        "Electricity production from coal sources (% of total)"
    ) |>
  select(renew_energy_consume, renew_electric_output, electric_prod_oil, 
         electric_prod_coal, cluster_2015) |>
  ggpairs(aes(color = cluster_2015),
          upper = list(continuous = "blank"),
          progress = FALSE) +
  theme(text = element_text(size = 8))
```

:::

In 1990:

- **Cluster 1 (red - Low-Moderately Sustainable)**  
  - Relatively high oil-based electricity production, with low renewable energy consumption and low renewable electricity output.
  - These countries lead in heavy reliance on oil for producing electricity while also having limited renewable resources.

- **Cluster 2 (green - Most Sustainable)**  
  - Higher levels of renewable energy consumption and electricity output, with low oil and coal electricity production.
  - Countries in this group tend to be ahead of the curve on sustainability. These are likely to include nations with strong environmental policies or access to abundant natural resources like hydropower. Think of places like Norway or Costa Rica.

- **Cluster 3 (blue - Least Sustainable)**  
  - Relatively high coal-based electricity production, with moderate renewable energy consumption.  
  - These countries still rely heavily on coal for electricity. Economic or political barriers might make renewable investment difficult—or fossil fuels may simply be more affordable or accessible in these regions.

- **Cluster 4 (purple - Moderately Sustainable)**  
  - Low to moderate values across all four indicators. Moderate coal usage but low in renewable energy usage. 
  - These countries have no extreme values in either fossil fuel use or renewable usage.

**Ranking from least to most sustainable groups in 1990**: 3, 1, 4, 2

In 2015:

- **Cluster 1**  
  - Remains the second least sustainable due to oil-based production, although with some improvement in renewable energy consumption.

- **Cluster 2**  
  - Remains the most sustainable. Clear leader in sustainability metrics.

- **Cluster 3**  
  - Lower values across all four indicators.  
  - Now the moderately sustainable group.

- **Cluster 4**  
  - Higher coal production than the other clusters. Moderate renewable energy and output.  
  - Least sustainable due to high coal-based production.

**Ranking from least to most sustainable groups in 2015**: 4, 1, 3, 2

### Summary tables of sustainability clusters
```{r}
# Add cluster labels back to the original indicator data
sustain1990_summary <- sustainability1990 |>
  inner_join(sustain_1990_c4 |> select(`Country Name`, cluster_1990), by = "Country Name")

sustain2015_summary <- sustainability2015 |>
  inner_join(sustain_2015_c4 |> select(`Country Name`, cluster_2015), by = "Country Name")

# 1990 Summary: Count + means
summary_1990 <- sustain1990_summary |>
  group_by(cluster_1990) |>
  summarise(
    Countries = n(),
    `Avg Renewable Energy Consumption (%)` = mean(`Renewable energy consumption (% of total final energy consumption)`),
    `Avg Renewable Electricity Output (%)` = mean(`Renewable electricity output (% of total electricity output)`),
    `Avg Oil Production Share (%)` = mean(`Electricity production from oil sources (% of total)`),
    `Avg Coal Production Share (%)` = mean(`Electricity production from coal sources (% of total)`)
  )

# 2015 Summary: Count + means
summary_2015 <- sustain2015_summary |>
  group_by(cluster_2015) |>
  summarise(
    Countries = n(),
    `Avg Renewable Energy Consumption (%)` = mean(`Renewable energy consumption (% of total final energy consumption)`),
    `Avg Renewable Electricity Output (%)` = mean(`Renewable electricity output (% of total electricity output)`),
    `Avg Oil Production Share (%)` = mean(`Electricity production from oil sources (% of total)`),
    `Avg Coal Production Share (%)` = mean(`Electricity production from coal sources (% of total)`)
  )
```

::: {.panel-tabset}

# 1990
```{r}
summary_1990 |>
  rename(`Cluster Group (1990)` = cluster_1990) |>
  kbl(booktabs = TRUE, digits = 2, align = "l") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

```

# 2015
```{r}
summary_2015 |>
  rename(`Cluster Group (2015)` = cluster_2015) |>
  kbl(booktabs = TRUE, digits = 2, align = "l") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

```
:::

From the table above, the number of countries classified as ‘least sustainable’ increased from 32 in 1990 to 48 in 2015, indicating a significant expansion of this group over time. Concurrently, six additional countries joined the ‘most sustainable’ cluster by 2015. Notably, countries within the same sustainability category often differ substantially in both geographic location and economic development. For instance, in 2015, the ‘most sustainable’ group included not only high-income nations such as Canada and the Nordic countries (Denmark, Norway, Sweden, and Finland) but also emerging and lower-income economies, including several Sub-Saharan African countries and Brazil. This suggests that sustainability outcomes are not strictly determined by income/GDP or geography, but may reflect differing national priorities, governance structures, or other circumstances.

## Sustainability cluster map

In 1990, The United States was a different sustainability landscape than in recent times and this world map visualizes this progress: It was among one of the least sustainable nations with higher coal-based electricity production as a share of its total electricity productionn and less renewable energy consumption and output. By 2015, the U.S. advanced to a moderate sustainablity level by more than doubling its renewable energy consumption and also increasing its renewable electricity output, while also reducing its share of total electricity production from both oil and coal sources. The story of environmental sustainability progress looks different for all 185 countries with publicly available data on the four energy indicators. Explore the choropleth world map below to investigate how the country of your interest has fared in the race towards a cleaner future.

::: {.panel-tabset}

### 1990 Map
```{r}
library(plotly)

# Prepare your sustainability + cluster data (already renamed):
data_1990 <- sustainability1990 |>
  left_join(clusters_1990, by = c("Country Name")) |>
  rename(
    `Renewable Energy (%)` = `Renewable energy consumption (% of total final energy consumption)`,
    `Renewable Electricity (%)` = `Renewable electricity output (% of total electricity output)`,
    `Oil Electricity (%)` = `Electricity production from oil sources (% of total)`,
    `Coal Electricity (%)` = `Electricity production from coal sources (% of total)`
  ) |>
  mutate(
    cluster_label = factor(cluster_1990,
                           levels = c(3, 1, 4, 2),
                           labels = c("Least Sustainable", "Low-Moderate", "Moderate", "Most Sustainable")),
    hover_text = paste0(
      "<b>", `Country Name`, "</b><br>",
      "Cluster: ", cluster_label, "<br>",
      "Renewable Energy: ", round(`Renewable Energy (%)`, 1), "%<br>",
      "Renewable Electricity: ", round(`Renewable Electricity (%)`, 1), "%<br>",
      "Oil Electricity: ", round(`Oil Electricity (%)`, 1), "%<br>",
      "Coal Electricity: ", round(`Coal Electricity (%)`, 1), "%"
    )
  )

# Map using ISO country codes
plot_1990 <- plot_geo(data_1990) |>
  add_trace(
    locations = ~`Country Code`,
    locationmode = "ISO-3",
    z = ~as.numeric(factor(cluster_1990, levels = c(3, 1, 4, 2))),
    colorscale = list(c(0, "#FF0000"), c(0.33, "#FF9900"), c(0.66, "#FFFF00"), c(1, "#33CC33")),  # red to green
    colorbar = list(
      title = "Cluster",
      tickvals = c(1, 2, 3, 4),
      ticktext = c("Least Sustainable", "Low-Moderate", "Moderate", "Most Sustainable"),
      xanchor = "center",
      orientation = 'h',
      x = 0.5,
      y = 0
    ),
    text = ~hover_text,
    hoverinfo = "text",
    showscale = TRUE
  ) |>
  layout(
    title = "Sustainability Clusters and Indicators in 1990",
    geo = list(showframe = TRUE, showcoastlines = TRUE)
  )
plot_1990

```

### 2015 Map
```{r}
# Prep 2015 data
data_2015 <- sustainability2015 |>
  left_join(clusters_2015, by = c("Country Name")) |>
  rename(
    `Renewable Energy (%)` = `Renewable energy consumption (% of total final energy consumption)`,
    `Renewable Electricity (%)` = `Renewable electricity output (% of total electricity output)`,
    `Oil Electricity (%)` = `Electricity production from oil sources (% of total)`,
    `Coal Electricity (%)` = `Electricity production from coal sources (% of total)`
  ) |>
  mutate(
    cluster_label = factor(cluster_2015,
                           levels = c(4, 1, 3, 2),
                           labels = c("Least Sustainable", "Low-Moderate", "Moderate", "Most Sustainable")),
    hover_text = paste0(
      "<b>", `Country Name`, "</b><br>",
      "Cluster: ", cluster_label, "<br>",
      "Renewable Energy: ", round(`Renewable Energy (%)`, 1), "%<br>",
      "Renewable Electricity: ", round(`Renewable Electricity (%)`, 1), "%<br>",
      "Oil Electricity: ", round(`Oil Electricity (%)`, 1), "%<br>",
      "Coal Electricity: ", round(`Coal Electricity (%)`, 1), "%"
    )
  )

# Map (2015)

plot_2015 <- plot_geo(data_2015) |>
  add_trace(
    locations = ~`Country Code`,
    locationmode = "ISO-3",
    z = ~as.numeric(factor(cluster_2015, levels = c(4, 1, 3, 2))),
    colorscale = list(c(0, "#FF0000"), c(0.33, "#FF9900"), c(0.66, "#FFFF00"), c(1, "#33CC33")),  # red to green
    colorbar = list(
      title = "Cluster",
      tickvals = c(1, 2, 3, 4),
      ticktext = c("Least Sustainable", "Low-Moderate", "Moderate", "Most Sustainable"),
      xanchor = "center",
      orientation = 'h',
      x = 0.5,
      y = 0
    ),
    text = ~hover_text,
    hoverinfo = "text",
    showscale = TRUE
  ) |>
  layout(
    title = "Sustainability Clusters and Indicators in 2015",
    geo = list(showframe = TRUE, showcoastlines = TRUE)
  )

plot_2015

```

:::
## Discussion

## References
* Zhu, Hao. “Create Awesome HTML Table with Knitr::Kable and Kableextra.” Cran, 23 Jan. 2024, https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html.
* “Hexadecimal HTML Color Codes.” Hexadecimal HTML Color Codes, https://html-color-codes.com/. 
* Sievert, Carson. “Interactive Web-Based Data Visualization with R, Plotly, and Shiny.” 4 Maps, 19 Dec. 2019, https://plotly-r.com/maps.

